---
title: "Creating New Scoring Criterion"
author: "Melat Ghebreselassie and Joe Brown"
date: "2024-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(matilda)
library(tidyverse)
```
## Background

The `matilda` package gives the user the power to create their own criterion to score model runs using `new_criterion`. This function takes data provided by the user and helps to build an object that can be used to weight `hector` runs.

In this example, we will create a criterion object that uses ocean carbon uptake data from the Global Carbon Project. The ocean sink data from GCP is estimated from the average of 10 ocean biogeochemistry models and 7 CO2 flux data products that reproduce the ocean sink since the 1990s. Prior to 1990, the ocean sink estimate is calculated from the multi-model mean of the 10 global ocean biogeochemistry models. More detailed information about the origin of this dat can be found in [Friedlingstein et al. 2023] (https://essd.copernicus.org/articles/15/5301/2023/#section6).

## Importing data 
We accessed the ocean uptake data from the Global Carbon Project. It needs some editing before it can be used for the criterion.
```{r, importing and editing ocean carbon data}

# Importing data from local repository
gcp_data <- read.csv("data/gcp_data.csv")

# Data munging for ocean C in useful year range
ocean_uptake_data <- gcp_data %>% 
  select(Year, ocean_sink) %>% # select the year and ocean_sink columns
  rename(value = ocean_sink, # rename columns if needed 
         year = Year) %>% 
  filter(year > 1849) %>% # can filter to include what year range you want. Here I show filtering 1850-present 
  na.omit()

```

This is how our data looks currently:
```{r, print head of the data, echo=FALSE}
head(ocean_uptake_data)
```

## Create New Criterion

Now we can use this data to create a new scoring criterion to weight `matilda` ensembles.

```{r, creating new criterion}

# Use new_criterion to create new scoring criterion
criterion_ocean_uptake <- new_criterion(var = OCEAN_UPTAKE(), # name of the variable in the model output used for comparison
                                        years = ocean_uptake_data$year, # year range of the scoring data
                                        obs_values = ocean_uptake_data$value # values used for scoring
                                        ) 

```

This code creates a new object that has an attribute class `criterion` and can be used as a criterion input when scoring `matilda` ensembles.

```{r, printing class information}
class(criterion_ocean_uptake)
```
When we print the object we see the criterion summarized.

```{r, printing criterion name}

print(criterion_ocean_uptake)

```

You are now poised to use this criterion to weight a `metilda` ensemble.

____

New temperature criterion:
```{r}
temp_data <- read.csv("data/temp_anomaly_annmean_hadcrut5.csv", stringsAsFactors = F)

temp_criterion_dat <- temp_data %>% 
  select(year, anomaly) %>% 
  rename(value = anomaly) %>% 
  filter(year > 1879) %>% 
  normalize_temperature(1961, 1990)


head(temp_criterion_dat)

```

```{r}
# Use new_criterion to create new scoring criterion
criterion_temp <- new_criterion(var = GMST(),# name of the variable in the model output used for comparison
                                years = temp_criterion_dat$year,
                                # year range of the scoring data
                                obs_values = temp_criterion_dat$adjusted_value) # values used for scoring

```


New CO2 criterion
```{r}
co2_data <- read.csv("data/historical_co2.csv", stringsAsFactors = F)

co2_criterion_data <- co2_data %>% 
  select(v.YEARS.GAS.., CO2) %>% 
  rename(year = v.YEARS.GAS..,
         value = CO2) %>% 
  filter(year > 1849)

head(co2_criterion_data)
```

```{r}
# Use new_criterion to create new scoring criterion
criterion_co2 <- new_criterion(var = CONCENTRATIONS_CO2(),# name of the variable in the model output used for comparison
                                years = co2_criterion_data$year,
                                # year range of the scoring data
                                obs_values = co2_criterion_data$value) # values used for scoring
```

