---
title: "Running Matilda Simulation"
author: "Joe Brown"
date: "2024-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal

The goal of this script is to run `matilda` to create a 10,000 member ensemble for each SSP scenario:

- `SSP1-1.9`

- `SSP1-2.6`

- `SSP2-4.5`

- `SSP3-7.0`

- `SSP5-8.5`

Compare these simulations against FaIR and MAGICC results reported in **Table 7.SM.4** of the [IPCC report] (https://www.ipcc.ch/report/ar6/wg1/downloads/report/IPCC_AR6_WGI_FGD_Chapter07_SM.pdf).

Temperature results will include `global_tas` normalized to 1995-2014 reference period. 

Ensembles will be weighted using `gmst`, `co2 concentrations`, and `ocean carbon uptake`. A goal here is to also adjust scoring influence to determine what gets the `matilda` ensemble closest to the values reported by FaIR and MAGICC.

Metrics will be computed for 3 different periods: `2021:2040`; `2041:2060`; and `2081:2100`.

Attempt to build paralleled `matilda` analysis into a function. 

#### Packages

```{r, load packages}
library(matilda)
library(parallel)
library(tidyverse)
```

#### Load ini list

```{r, import ini files}
# locate the ini directory
ini_dir <- paste0(system.file("input", package = "hector"), "/")

# read in ini files into a list
ini_list <- list(ssp119 = paste0(ini_dir, "hector_ssp119.ini"),
                 ssp245 = paste0(ini_dir, "hector_ssp245.ini"),
                 ssp370 = paste0(ini_dir, "hector_ssp370.ini"),
                 ssp585 = paste0(ini_dir, "hector_ssp585.ini"))
```

#### Peturbed parameter set

```{r, build perturbed parameter data frame}
# Using generate_params() to build the perturbed parameter set

# set seed for replication
set.seed(123)

# set sample size (we will run with a smallish sample size now)
n = 10

# initiate a core using the first ini file in ini_list
params_core <- newcore(ini_list[[1]])

# produce parameter set 
params <- generate_params(core = params_core,
                          draws = n)
```

#### Split jobs for effective parallel

```{r, spliting params data frame into chunks}
# split params into chunks
param_chunks <- split(params, 1:5)
```

#### Run model

```{r, run the model}
# initializing a cluster
cl <- makeCluster(detectCores() - 1)

# Export required functions and objects to the cl cluster we just created
clusterExport(cl, c("param_chunks",
                    "ini_list",
                    "newcore",
                    "iterate_model"))

# run the model with parallel computing
result <- parLapply(cl, names(ini_list), function(scenario_name){
  
  # extract the scenario information from the ini_list 
  # using the scenario name
  scenario <- ini_list[[scenario_name]]
  
  # initialize model core for the current scenario
  core <- newcore(scenario, name = scenario_name)
  
  # run the model looping across param_chunks for the current core
  result_list <- lapply(param_chunks, function(chunk) {
    
    iterate_model(core = core, 
                  params = chunk, 
                  save_years = 1800:2100,
                  save_vars = c("global_tas", "gmst", "ocean_uptake", "CO2_concentration"))
  })
  
  ## This step ensures a correct run_numbers are added to each model run ##
  # Starting with the second data frame of the current scenario
  for (i in 2:length(result_list)) {
    
    # calculate the max value of the previous element in the result list
    max_run_number <- max(result_list[[i - 1]]$run_number)
    
    # Add the max value of the previous element to the run_number of the current 
    # element to get an updated run_number that is continuous from the previous element.
    result_list[[i]]$run_number <- result_list[[i]]$run_number + max_run_number
    
  }
  
  # bind 
  result <- do.call(rbind, result_list)
  
  return(result)
  
})

#close cluster
stopCluster(cl)

# bind results to create a data frame
results_df <- do.call(rbind, result)
```

#### Score Model Results

```{r, score matilda ensemble}
# loop through results in model_results to compute model weights for each 
score_list <- lapply(result, function(df){
  
  # use observed temperature to compute model weights
  score_sensitivity(df, mc_weights_name = "balanced_weight")
  
})
```

#### Normalize result

```{r}

```

#### Compute metrics

```{r}
metric_list <- list(
  short = new_metric(var = "global_tas", years = 2021:2040, op = median),
  mid = new_metric(var = "global_tas", years = 2041:2060, op = median),
  long = new_metric(var = "global_tas", years = 2081:2100, op = median))
```

```{r}
# loop through results and compute metrics for each
metric_result <- lapply(result, function(df){

  metric_caclulation <- lapply(metric_list, function(metric) {
  
  # calculate metrics using the metric we defined in the previous step
  metric_calc(df, metric)

  })
  
  return(metric_caclulation)
  
})
```

